---
title: "DS2003Midterm"
output: html_document
date: "2022-10-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(tidyverse)
```
### **Our chosen dataset examines over 200 students enrolled in a math course at a University in Portugal . Within our dataset is  a number of social, gender, and class variables with information that provide insight into the lives of these students. By examining this dataset we can explore the relationship between factors such as alcohol consumption, gender, age, romantic relationships, activites, absences, and even what your environment/family is like in order to gauge how each of these environmental factors can affect performance in school. The key variables within our dataset are- the number of past class failures (failure), the student’s age (age), going out with friends (goout-on a scale of 1-5), in a romantic relationship (romantic),  weekly study time (study time), internet access at home (internet), student’s sex (sex), and whether or not they wanted to pursue higher education (higher). While there were 33 variables examined in the study including those calculating the student’s grades throughout the semester, these eight variables were found to have the highest correlation to how well a student performed and are therefore our key variables.**

### **Question 1: Question 2: How does each variable correlate to the students’ final grade?**

### **Hypothesis: Alcohol consumption is strongly tied to one’s final grade**

```{r cars}
library(corrr)
library(corrplot)
Grades <- read_csv("D:/DS_2003/DS-2003/student-mat.csv")
view(Grades)
Grades[,c(1,2,4,5,6,9,10,11,12,16,17,18,19,20,21,22,23)] <- lapply(Grades[,c(1,2,4,5,6,9,10,11,12,16,17,18,19,20,21,22,23)], as.factor)
str(Grades)

drops <- c("reason","Fjob","Mjob","Fedu","Medu","address","reason","guardian","school","paid")
Grades = Grades[ , !(names(Grades) %in% drops)]
Grades[] <- lapply(Grades, as.integer)
str(Grades)
colnames(Grades) <- c("Sex","Age","Family Size","Parent Status","Travel Time","Study Time","Past Failed Classes","School Education Support","Family Education Support","Extra-Curs","Attended Nursery","Higher Ed","Internet","Romantic","Family Relationship","Free Time","Going Out","Workday Alc Consumption","Weekend Alc Consumption","Health","Absences","G1","G2","G3")
view(Grades)
Grades_cor <- cor(Grades[1:21], Grades[22:24])
Grades_cor <- Grades_cor[order(Grades_cor[1:21]),]
Grades_cor <- as.data.frame(Grades_cor)
Grades_cor["rownames"] <- rownames(Grades_cor)
view(Grades_cor)
Grades_cor <- Grades_cor %>% 
  mutate(interval  = case_when(G3 < 0 ~ 'below', G3 >= 0 ~ 'above'))
Grades_cor <- Grades_cor %>%
  arrange(desc(G3))
view(Grades_cor)
Grades_cor %>%
  ggplot(aes(x = reorder(rownames,G3), y = G3, fill = interval)) + 
  geom_bar(stat = "identity") +
  ylab("r Correlation value with final grade") +
  xlab("Variable") +
  theme(text = element_text(size=12), axis.text.x = element_text(angle=50,hjust=1, size = 10),legend.position="none",axis.line = element_line(colour = "black"),panel.background = element_rect(fill = "white")) + geom_text(aes(x = rownames, y = G3, label = round(G3,3)),
    position = position_dodge(width = 1),
    vjust = -0.3, size = 2.5) +
  ggtitle("Correlation: R vs. Variables")
  #png(filename = "Bars.png")
```
### **Conclusion: Alcohol consumption is not that strongly correlated with one's final grade; number of past failed clases, age, and wanting to go onto higher education have the largest correlation values. As seen in the graph, # of past failed classes is most negatively correlated and age and the desire for higher education are the two most positively correlated values. Although we can't make definite conclusions about the data from this trend alone, it does point us into a direction for further investigation. But as we can see from the graph, the amount of alcohol consumption has little correlated with one's finale grade, the opposite of our hypothesis.**

```{r}
stud_alc <- read.csv("D:/DS_2003/DS-2003/student-mat.csv")
```

## **Question:** Does the level of weekend alcohol consumption in students relate to their final grades?

### **Hypothesis:** We believe that grades will decrease in students with higher weekend alcohol consumption levels.

```{r}
#Wrangling data to make a tibble with the average grade percent for each level of alcohol consumption
stud_alc_summarized <- stud_alc %>% mutate(G1 = (G1/20)*100) %>% group_by(as.factor(Walc)) %>% summarise(mean = round(mean(G1), 2), sd = sd(G1)) %>% rename('Weekend_Alcohol_Consumption' = 'as.factor(Walc)') %>% mutate(pct = '%') %>% unite(bar_label, c(mean, pct), sep = '', remove = FALSE)
stud_alc_summarized
#Creating the visualization using ggplot
ggplot(stud_alc_summarized, aes(x = Weekend_Alcohol_Consumption, y = mean, fill = 'pink')) + geom_bar(stat = 'identity', position = 'dodge') + theme_classic() + labs(title = "Average First Period Grade by Weekend Alcohol Consumption", caption = "Error bars represent +/-1 standard deviation from the average") + xlab("Weekend Alcohol Consumption") + ylab("First Period Grade (%)") +
  geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=.2,
                 position=position_dodge(.9)) + theme(legend.position = 'none') +geom_text(aes(label = bar_label), vjust = 10, colour = "black")
```

### **Findings:** 
From the graph, we see that the grades of students worsen as alcohol consumption increases, especially as students cross a threshold from the 1-3 range to the 4-5 range. However, we cannot say that alcohol consumption causes grades to fall for two major reasons. First, a confounding variable could exist that causes both variables to vary with each other. For instance, perhaps course for older students have typically have lower grades across the class, and these students only drink more because they have better access to alcohol as legal students. Secondly, we would not be able to know which way the relationship between the two variables, so we cannot assume causality. It would be plausible for students who receive worse grades to turn to alcohol as a way to drown their sorrows away.

Additionally, although a five percent difference would be significant for many students (although, these students would average F's using UVa's grading scale), we cannot say there is a significant difference between these grades. We did not perform an statistical tests such as a ANOVA test to check whether the difference should be attributed to chance instead. Looking at the error bars (1 standard deviation from the averages in each direction or ~68% of the sample assuming normality), there are many students who seldom drink alcohol and still score below the average of those who drink alcohol heavily and many students who drink large amounts of alcohol and perform better in the classroom than the average of those students who drink small amounts.

Altogether, although we see a trend in the relationship between alcohol consumption and first period grade, it appears to follow more of a piece wise pattern than a linear or exponential pattern, and we cannot make any definitive conclusions.
```
